---
title: "ITS DADA2"
output: html_document
date: "2025-10-27"
---

#### ITS specific workflow by https://benjjneb.github.io/dada2/ITS_workflow.html
```{r}
library(dada2)
library(openxlsx)
library(vegan)
library(tidyverse)
library(ShortRead)
library(Biostrings)
```

#### ITS ANALISI
```{r}
path <- "~/Downloads/FEM_AMPLICON_RUN011-20260323T143521Z-1-001/FEM_AMPLICON_RUN011/ITS_folder/trimmed"
setwd(path)
list.files(path)

fnFs <- sort(list.files(path, pattern="_R1.trimmed.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2.trimmed.fastq.gz", full.names = TRUE))

sample.names <- sapply(strsplit(basename(fnFs), "_"), '[', 1)
#sample.names = gsub("-", "_", sample.names)
#sample.names = gsub("NBFC_16S_","", sample.names)

#view sample names
sample.names
```

```{r}
plotQualityProfile(fnFs[1:4])
plotQualityProfile(fnRs[1:4])
```
##### Quality filtering #ITS version, with maxErr
```{r}
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, maxN = 0, maxEE = c(2, 2), truncQ = 2,
    minLen = 50, rm.phix = TRUE, compress = TRUE, multithread = TRUE)  
head(out)

```


```{r}
#The output shows the fraction of reads discarded. The quality of the filtered filed can be also cheked:

plotQualityProfile(filtFs[1:4])
plotQualityProfile(filtRs[1:4])
```
##### Learning error rates
```{r}
#DADA2 removes all sequencing errors to reveal the members of the sequenced community.The error profiles are used in a successive step to correct errors.
errF <- learnErrors(filtFs, multithread = TRUE)
errR <- learnErrors(filtRs, multithread = TRUE)

plotErrors(errF, nominalQ=TRUE)
plotErrors(errR, nominalQ=TRUE)
```


##### Sample inference step
```{r}
# The sample inference algorithm is applied to the filtered and trimmed sequence data, with the aim to infer true biological sequences. This is done by incorporating the quality profiles and abundances of each unique sequence, deciding if sequences are “true” (of biological origin), or spurious
dadaFs <- dada(filtFs, err=errF, pool=FALSE, multithread=TRUE)
dadaRs <- dada(filtRs, err=errR, pool=FALSE, multithread=TRUE)
```

##### Merge paired reads
```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
```

##### Construct amplicon sequence variants (AVs) table
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)

plot(table(nchar(getSequences(seqtab))), xlab="Reads R1+R2 merged length") #result:a natural, multimodal distribution centered around 230bp to 300bp, with a "long tail" extending past 400bp. This represents the actual diversity of fungal ITS1/ITS2 regions in your samples
```

#### Remove chimeras 
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim)/sum(seqtab)*100 # number of reads after removing chimeras
```
As a further computational check, it is worth to know how many reads were discarded at
various points of the pipeline:
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers,
getN), rowSums(seqtab.nochim))
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
```
##### assign Taxonomy on server
1. prep sequences
```{r}
seqs <- colnames(seqtab.nochim)
length(seqs)
#saveRDS(seqs, file = "/Users/macbook/Documents/Work/FEM/FEM/NBFC/Tassonomia/seqs_ITS.rds")
```
##### Assign Taxonomy on Server
```{r}
unite.ref <- "~/Downloads/sh_general_release_19.02.2025/sh_general_release_dynamic_19.02.2025.fasta" # CHANGE ME to location on your machine
taxa <- assignTaxonomy(seqtab.nochim, unite.ref, multithread = TRUE, tryRC = TRUE)
```


```{r}
# save sequences with original headers
#write.csv(t(seqtab.nochim), "/Users/macbook/Documents/Work/FEM/FEM/NBFC/Tassonomia/assigned_reads/ITS_seqtab-nochim.csv", quote=FALSE)

# simplify names to sequence headers(seq1, seq2...seq100...)
# adapted from: https://github.com/benjjneb/dada2/issues/655
seqs <- colnames(seqtab.nochim)
SeqName <- vector(dim(seqtab.nochim)[2], mode="character")
SeqName_ft <- vector(dim(seqtab.nochim)[2], mode="character")
for (i in 1:dim(seqtab.nochim)[2]) {
 SeqName[i] <- paste("seq", i, sep="")
 SeqName_ft[i] <- paste(">seq", i, sep="")
}

# write sequences in a FASTA file
fastaseqs_ft <- rbind(SeqName_ft, seqs)
#write(fastaseqs_ft, "/Users/macbook/Documents/Work/FEM/FEM/NBFC/Tassonomia/assigned_reads/ITS_fastaseqs.fasta")
# write sequences in a FASTA file (tabula)
fastaseqs <- cbind(SeqName, seqs)
#write.csv(fastaseqs, paste0(pathtab,"fastaseqs.csv"), quote=FALSE, row.names = FALSE)

# write count table
seqtab.nochim.t <- t(seqtab.nochim)
row.names(seqtab.nochim.t) <- SeqName
seqtab.nochim.t <- tibble::rownames_to_column(as.data.frame(seqtab.nochim.t), "SeqName")
#write.csv(seqtab.nochim.t, "/Users/macbook/Documents/Work/FEM/FEM/NBFC/Tassonomia/assigned_reads/counts.csv", quote=FALSE, row.names = FALSE)

# write taxonomy table
row.names(taxa) <- SeqName
taxatibble <- tibble::rownames_to_column(as.data.frame(taxa), "SeqName")
#write.csv(taxatibble, "/Users/macbook/Documents/Work/FEM/FEM/NBFC/Tassonomia/assigned_reads/ITS_taxa.csv", quote=FALSE, row.names = FALSE)
# Join results and save in spreadsheet (excel) format
tax_counts <- left_join(taxatibble, seqtab.nochim.t, by = "SeqName", keep = TRUE)
tax_counts_fasta <- left_join(tax_counts, as.data.frame(fastaseqs), by = c("SeqName.x" =
"SeqName"), keep = TRUE)
 openxlsx::write.xlsx(tax_counts_fasta, file = "~/Downloads/FEM_AMPLICON_RUN011-20260323T143521Z-1-001/tax_counts_fasta_its_run011.xlsx",overwrite = TRUE, asTable = FALSE, sheetName = "NBFC_ITS", firstRow = TRUE, zoom = 90, keepNA = TRUE)
```

#Fasta sequences for MUSCLEv5 alignment
```{r}
fastafile=read_excel("/Users/macbook/Documents/Work/FEM/FEM/NBFC/Tassonomia/assigned_reads/ITS_tax_counts_fasta.xlsx")
fastaseqs=fastafile[,c(130:131)]
names(fastaseqs)[1]="SeqName"
# Function to write FASTA
write_fasta <- function(names, seqs, file) {
  con <- file(file, "w")
  for (i in seq_along(names)) {
    writeLines(paste0(">", names[i]), con)
    writeLines(seqs[i], con)
  }
  close(con)
}

# Save as FASTA
write_fasta(fastaseqs$SeqName, fastaseqs$seqs, "/Users/macbook/Documents/Work/FEM/FEM/NBFC/Tassonomia/assigned_reads/its.fasta")
```

```{r}
# VEGAN
library(vegan)
counts <-read.csv(file="~/Documents/Work/FEM/FEM/NBFC/Tassonomia/assigned_reads/ITS_seqtab-nochim.csv", header=T, row.names=1, check.names=FALSE)
head(counts)
counts.t <- t(counts)
names(counts)=gsub("_L001_R1.trimmed.fastq.gz","",names(counts))
names(counts)=gsub("NBFC-ITS-","",names(counts))
counts2 <- t(counts)^0.25
bc <- vegdist(counts2, method="bray")
plot(hclust(bc))
```

```{r}
# PHYLOSEQ ANALYSIS
library(phyloseq)
library(ggplot2)
metadata <- read.csv(file="/Users/macbook/Documents/Work/FEM/FEM/NBFC/Metadati/composizione_chimica.csv", header=T, row.names=1, sep = ";")
nbfc_its <- phyloseq(otu_table(counts, taxa_are_rows=TRUE), sample_data(metadata), tax_table(taxatibble))
```

```{r}
nbfc_its <- phyloseq(otu_table(counts, taxa_are_rows=TRUE), taxa_names(taxatibble))

```

